---
title: "Exploratory Analysis of School Shootings"
author: "Significantly Different"
date: "March 29, 2019"
output: github_document
---

## Section 1. Introduction

Gun control has been an issue of huge debate in the United States: from individuals claiming that the possession of guns is necessary to protect one’s own safety to others who claim that it’s a “right” bestowed by the constitution, “gun control” seems to imply different things to everyone. With a huge market of gun sales being run by capitalism, it doesn’t seem like any number of casualties caused by gun violence will be sufficient to make up for the monetary losses a ban on guns would have on our capitalist market — which is to some extent why we still have gun control issues going on in the United States. Such issue of gun violence has unfortunately been spilling into our schools, and school shootings have become the issue of concern for teachers, school districts, and families in hopes of never having to face such an incident. The lack of control and support the government and community provided towards this issue has unfortunately lead to schools carrying the burden of coming up with preventative measures for school shootings to ensure students’ safety — a high school in South Florida, for example, has brought up and implemented the idea of a bullet-proof shelter to protect children from harm, and another school in Texas asked the students to wear a clear backpack to school. Despite such efforts, school shootings still break out, leading researchers and school officials to increasingly work more to figure out the best ways to prevent school shootings from occurring.

The aim of the current project is to look into data of school shootings from the past to evaluate what factors correlate to higher number of casualties in school shootings, in hopes that this analysis will help lead to preventive measures.

Throughout this project, our team will look into the ```schools-shootings-data.csv``` as provided by the Washington Post. The data consists of any act of gunfire at schools (primary, secondary) that occurred during school hours, or immediately before or after hours, since the Columbine High massacre on April 20, 1999. As such, 221 incidences of school shootings have been included in the dataset. The data was collected by using “Nexis, news articles, open-source databases, law enforcement reports, information from school websites, and calls to schools and police departments”, as according to Washington Post. The variables included in this data are as follows: 

uid, nces_school_id, school_name, nces_district_id, district_name, date, school_year, year, time, day_of_week, city, state, school_type, enrollment, killed, injured, casualties, shooting_type, age_shooter1, gender_shooter1, race_ethnicity_shooter1, shooter_relationship1, shooter_deceased1, deceased_notes1, age_shooter2, gender_shooter2, race_ethnicity_shooter2, shooter_relationship2, shooter_deceased2, deceased_notes2, white, black, Hispanic, Asian, American_indian_alaska_native, Hawaiian_native_pacific_islander, two_or_more, resource_officer, weapon, weapon_source, lat, long, staffing, low_grade, high_grade, lunch, county, state_fips, county_fips, ulocale

More information on the description of these variables can be found under the codebook in the README file. 

## Section 2. Data analysis plan

For our analysis, we plan to use variables that seem likely to have an influence on the number of casualties. Our response (dependent) variable will be the number of casualties while the explanatory (independent) variables will range from 1 to 50, according to the variables we decide to include and that have an effect on the number of casualties, such as ```shooting_type```, ```age_shooter1```, and racial demographics of the school, as determined by the variables ```white```, ```black```, ```hispanic```, ```asian```, ```american_indian_alaska_native```, ```hawaiian_native_pacific_islander```, and ```two_or_more```. We might use comparison groups that relate to income, race or type of high school. However, this is not finalized. 

Now we perform some preliminary exploratory data analysis by showing summary statistics and visualizations:


```{r load-packages, echo = FALSE, message = FALSE}
library(tidyverse)
```

```{r load-data, echo = FALSE, message = FALSE}
school_shootings <- read_csv("../data/school-shootings-data.csv")
```

We start by observing the distribution of casualties (our response variable):

```{r summ-stats, echo = FALSE}
school_shootings %>%
  filter(!is.na(casualties)) %>%
  ggplot(mapping = aes(x = casualties)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of casualties") +
  xlab("Casualties") +
  ylab("Frequency")
```
This will be useful as we would want to see what the median/mean number of casualties is as well as the general shape of our distribution. It looks like it is skewed to the right, which means we would look at the median:

```{r median-casualties, echo=FALSE}
school_shootings %>%
  summarise(median = median(casualties))
```
The median of number of casualties is 1. 

Next, we see the distribution of school staffing:

```{r staff-viz, echo=FALSE}
school_shootings %>%
  filter(!is.na(staffing)) %>%
  ggplot(mapping = aes(x = staffing)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Distribution of number of full-time teachers", subtitle = "At each shooting") +
  xlab("Amount of teachers") +
  ylab("Frequency")
```
Because this distribution is slightly skewed, we stick to the median. 

```{r mean-staff-num, echo = FALSE}
school_shootings %>%
  filter(!is.na(staffing)) %>%
  summarise(median_staff_num = median(staffing))
```
The median number of staff at schools is 54.53. This variable is useful because much of media coverage around school shooters mentions mental health and social support. It would be interesting to see if a higher number of staffing is correlated to lower or higher number of casualties.

We can also rank states by the number of school shootings that have occurred there:

```{r state-rank, echo=FALSE}
school_shootings %>%
  filter(!is.na(state)) %>%
  group_by(state) %>%
  count() %>%
  arrange(desc(n))
```
This variable is useful to observe the effect of states' gun control laws on the number of casualties or shootings happening. 

One more thing that would be interesting to see is the distribution of the amount of students who qualify for free or reduced lunch.

```{r lunch-viz, echo=FALSE}
school_shootings %>%
  filter(!is.na(lunch)) %>%
  ggplot(mapping = aes(x = lunch)) +
  geom_histogram()
```

However, this might be misleading given that this is a raw number, whereas a proportion would be more useful.


## Section 3. Data

Our data was not put in the ```/data``` folder since it started giving us issues (the R chunk would not find the data because there is a project folder under the larger project folder). Our codebook and dimensions were put in the README file.
