---
title: "Exploratory Analysis of School Shootings"
author: "Significantly Different"
date: "March 29, 2019"
output: github_document
---

## Section 1. Introduction

Gun control has been an issue of huge debate in the United States: from individuals claiming that the possession of guns is necessary to protect one’s own safety to others who claim that it’s a “right” bestowed by the constitution, “gun control” seems to imply different things to everyone. With a huge market of gun sales being run by capitalism, it doesn’t seem like any number of casualties caused by gun violence would be sufficient to make up for the number of dollars a ban on guns would have on our capitalist market—which is to some extent why we still have gun control issues going on in the United States. Such issue of gun violence has unfortunately been spilling into our schools, and school shootings has become the issue of concern for teachers, school districts, and families in hopes of never having to face such an incident. The lack of control and support the government and community provided towards this issue has unfortunately lead to schools carrying the burden of coming up with preventive measures for school shootings to ensure students’ safety—a high school in South Florida, for example, has brought up and implemented the idea of a bullet-proof shelter to protect children from harm, and another school in Texas asked the students to wear a clear backpack to school. Despite such efforts, school shootings still breakout, leading researchers and school officials to increasingly work more to figure out the best ways to prevent school shootings from occurring.

The aim of the current project is to look into data of school shootings from the past to evaluate what factors correlate to higher number of casualties in school shootings, in hopes that this analysis would help lead to preventive measures.

Throughout this project, our team will look into the ```schools-shootings-data.csv``` as provided by the Washington Post. The data consists of any act of gunfire at schools (primary, secondary) that occurred during school hours, or immediately before or after hours, since the Columbine High massacre on April 20, 1999. As such, 221 incidences of school shootings have been included in the dataset. The data was collected by using “Nexis, news articles, open-source databases, law enforcement reports, information from school websites, and calls to schools and police departments”, as according to Washington Post. The variables included in this data are as follows: 

uid, nces_school_id, school_name, nces_district_id, district_name, date, school_year, year, time, day_of_week, city, state, school_type, enrollment, killed, injured, casualties, shooting_type, age_shooter1, gender_shooter1, race_ethnicity_shooter1, shooter_relationship1, shooter_deceased1, deceased_notes1, age_shooter2, gender_shooter2, race_ethnicity_shooter2, shooter_relationship2, shooter_deceased2, deceased_notes2, white, black, Hispanic, Asian, American_indian_alaska_native, Hawaiian_native_pacific_islander, two_or_more, resource_officer, weapon, weapon_source, lat, long, staffing, low_grade, high_grade, lunch, county, state_fips, county_fips, ulocale

More information on the description of these variables can be found under the codebook in the README file. 

## Section 2. Data analysis plan

```{r load-packages}
library(tidyverse)

```

```{r load-data}
school_shootings <- read_csv("../project/school-shootings-data.csv")
```

```{r summ-stats}
school_shootings %>%
  filter(!is.na(casualties)) %>%
  ggplot(mapping = aes(x = casualties)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of casualties") +
  xlab("Casualties") +
  ylab("Frequency")
```


```{r staff-viz}
school_shootings %>%
  filter(!is.na(staffing)) %>%
  ggplot(mapping = aes(x = staffing)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Distribution of number of full-time teachers", subtitle = "At each shooting") +
  xlab("Amount of teachers") +
  ylab("Frequency")
```

```{r mean-staff-num}
school_shootings %>%
  filter(!is.na(staffing)) %>%
  summarise(median_staff_num = median(staffing))
```

We can also rank states by the number of school shootings that have occurred there:

```{r state-rank}
school_shootings %>%
  filter(!is.na(state)) %>%
  group_by(state) %>%
  count() %>%
  arrange(desc(n))
```

One more thing that would be interesting to see is the distribution of the amount of students who qualify for free or reduced lunch.

```{r lunch-viz}
school_shootings %>%
  filter(!is.na(lunch)) %>%
  ggplot(mapping = aes(x = lunch)) +
  geom_histogram()
```

However, this might be misleading given that this is a raw number, whereas a proportion would be more useful.


## Section 3. Data

