---
title: "Exploratory Analysis of School Shootings"
author: "Significantly Different"
date: "March 29, 2018"
output: github_document
---

## Section 1. Introduction

Gun control has been an issue of huge debate in the United States: from individuals claiming that the possession of guns is necessary to protect one’s own safety to others who claim that it’s a “right” bestowed by the constitution, gun control seems to imply different things to different people. With a huge market of gun sales being run by capitalism, it doesn’t seem like any number of casualties caused by gun violence will be sufficient to make up for the monetary loss a ban on guns would have on our capitalist market — which is to some extent why we still have gun control issues going on in the United States. Such issue of gun violence has unfortunately been spilling into our schools, and school shootings have become the issue of concern for teachers, school districts, and families in hopes of never having to face such an incident. The lack of control and support the government and community provided towards this issue has unfortunately lead to schools carrying the burden of coming up with preventative measures to ensure their students’ safety — a high school in South Florida, for example, has brought up and implemented the idea of a bullet-proof shelter to protect children from harm, and another school in Texas asked the students to wear a clear backpack to school. Despite such efforts, school shootings still break out, leading researchers and school officials to increasingly work more to figure out the best ways to prevent school shootings from occurring.

The aim of the current project is to look into data of school shootings from the past to evaluate what factors correlate to higher number of casualties in school shootings, in hopes that this analysis will help lead to preventive measures.

Throughout this project, our team will look into the ```schools-shootings-data.csv``` as provided by the Washington Post. The data consists of any act of gunfire at schools (primary, secondary) that occurred during school hours, or immediately before or after hours, since the Columbine High massacre on April 20, 1999. As such, 221 incidences of school shootings have been included in the dataset. The data was collected by using “Nexis, news articles, open-source databases, law enforcement reports, information from school websites, and calls to schools and police departments”, as according to Washington Post. The variables included in this data are as follows: 

uid, nces_school_id, school_name, nces_district_id, district_name, date, school_year, year, time, day_of_week, city, state, school_type, enrollment, killed, injured, casualties, shooting_type, age_shooter1, gender_shooter1, race_ethnicity_shooter1, shooter_relationship1, shooter_deceased1, deceased_notes1, age_shooter2, gender_shooter2, race_ethnicity_shooter2, shooter_relationship2, shooter_deceased2, deceased_notes2, white, black, Hispanic, Asian, American_indian_alaska_native, Hawaiian_native_pacific_islander, two_or_more, resource_officer, weapon, weapon_source, lat, long, staffing, low_grade, high_grade, lunch, county, state_fips, county_fips, ulocale

More information on the description of these variables can be found under the codebook in the ```data``` folder. 

## Section 2. Data Analysis



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-packages, message=FALSE}
library(tidyverse)
library(broom)
library(lubridate)

```

```{r load-data, message=FALSE}
school_shootings <- read_csv("../data/school-shootings-data.csv")
```

### Mutation of Variables

We begin by manipulating some of our variables in order to make them more useful. For example, many of our variables are raw and thus, converting them to percentages will be more useful and standardized. 

```{r mutation-races, echo=FALSE}
school_shootings <- school_shootings %>%
  mutate(white_percent = white/enrollment,
         black_percent = black/enrollment,
         hispanic_percent = hispanic/enrollment,
         asian_percent = asian/enrollment)
```

```{r lunch-percent, echo=FALSE}
school_shootings <- school_shootings %>%
  mutate(lunch_percent = lunch/enrollment)
```

```{r}
school_shootings <- school_shootings %>%
  mutate(time = hms::as.hms(time))

#school_shootings <- school_shootings %>%
#  mutate(time_day = case_when(
#    time >= hms::as.hms(6:00:00) & time < hms::as.hms(12:00:00) ~ "morning"
#  ))
```

We create the following variables:

<center>

```white_percent```
```black_percent```
```hispanic_percent```
```asian_percent```
```lunch_percent```

</center>

### Frequency of Annual School Shootings

```{r quan-shootings, fig.width=5, echo=FALSE}
school_shootings %>%
  group_by(year) %>%
  count() %>%
  ggplot(mapping = aes(x = year, y = n)) +
  geom_line() +
  labs(title = "Number of annual school shootings", subtitle = "1999 - 2018") +
  ylab("Frequency") +
  xlab("Year")

```

### Number of Casualties Across the Years

```{r freq-casualties, fig.width=6, echo=FALSE}
school_shootings %>%
  ggplot(mapping = aes(x = year, y = casualties)) +
  geom_point() +
  labs(title = "Casualties over the years 1999 - 2018") +
  ylab("Casualties") +
  xlab("Year")
```

### Distribution of Casualties

```{r dist-casualties, fig.width=5, echo=FALSE}
school_shootings %>%
  ggplot(mapping = aes(x = casualties)) +
  geom_bar() +
  labs(title = "Distribution of Casualties", subtitle = "School Shootings from 1999-2018") +
  xlab("Casualties") +
  ylab("Frequency")
```

### School with most casualties

The two schools with the most casualties (34) are Columbine High School and Marjory Stoneman Douglas High School. The shooting at Columbine High School took place in 1999 and the one at Douglas HS took place 19 years later in 2018. 

### Multivariate Regression Analysis

These are the variables that we will include in our model:

<center>

```enrollment``` 
```factor(shooting_type)```
```staffing```
```white_percent```
```black_percent```
```hispanic_percent```
```asian_percent```
```lunch_percent```
```factor(resource_officer)```
```factor(day_of_week)```

</center>

### Model Creation and Backwards Selection

First, we eliminate variables that will not be useful in our model, such as those used for logistical purposes. 

```{r sel-vars, echo=FALSE}
school_shootings <- school_shootings %>%
  select(school_name, 
         district_name, 
         date, 
         school_year, 
         year, 
         time, 
         day_of_week,
         lunch,
         city, 
         state, 
         school_type, 
         enrollment, 
         killed, 
         injured, 
         casualties, 
         shooting_type, 
         age_shooter1, 
         age_shooter2, 
         gender_shooter1, 
         gender_shooter2, 
         race_ethnicity_shooter1, 
         race_ethnicity_shooter2, 
         shooter_relationship1, 
         shooter_relationship2, 
         resource_officer, 
         weapon, 
         staffing, 
         white_percent, 
         black_percent, 
         hispanic_percent, 
         asian_percent, 
         lunch_percent)
```

### Full Model

We first create a full model, which includes all potential variables. The ```step()``` function automates the backwards selection process for us.

```casualties ~ enrollment + factor(shooting_type) + staffing + white_percent + black_percent + hispanic_percent + asian_percent + lunch_percent + factor(resource_officer) + factor(day_of_week)```

### Backwards Selection

```{r back-sel, results='hide', echo=FALSE}
mod_cas <- lm(casualties ~ 
              enrollment +
              factor(shooting_type) +
              staffing + 
              white_percent + 
              black_percent +
              hispanic_percent +
              asian_percent +
              lunch_percent +
              factor(resource_officer) +
              factor(day_of_week),
              #(enrollment * staffing) +
              #(lunch_percent * factor(resource_officer)),
              data = school_shootings)

#glance(mod_cas)$adj.r.squared

perf_model <- step(mod_cas, direction = "backward")

perf_model %>%
  tidy() %>%
  select(term, estimate)

glance(perf_model)$adj.r.squared

```

The perfect model is shown below:

```casualties ~ enrollment + staffing + lunch_percent + factor(resource_officer) + factor(shooting_type)```

Its adjusted r-squared value is 0.1699588.

