---
title: "Exploratory Analysis of School Shootings"
author: "Significantly Different"
date: "March 29, 2018"
output: github_document
---

## Section 1. Introduction

Gun control has been an issue of huge debate in the United States: from individuals claiming that the possession of guns is necessary to protect one’s own safety to others who claim that it’s a “right” bestowed by the constitution, gun control seems to imply different things to different people. With a huge market of gun sales being run by capitalism, it doesn’t seem like any number of casualties caused by gun violence will be sufficient to make up for the monetary loss a ban on guns would have on our capitalist market — which is to some extent why we still have gun control issues going on in the United States. Such issue of gun violence has unfortunately been spilling into our schools, and school shootings have become the issue of concern for teachers, school districts, and families in hopes of never having to face such an incident. The lack of control and support the government and community provided towards this issue has unfortunately lead to schools carrying the burden of coming up with preventative measures to ensure their students’ safety — a high school in South Florida, for example, has brought up and implemented the idea of a bullet-proof shelter to protect children from harm, and another school in Texas asked the students to wear a clear backpack to school. Despite such efforts, school shootings still break out, leading researchers and school officials to increasingly work more to figure out the best ways to prevent school shootings from occurring.

The primary focus of the current project is to look into the data of school shootings from the past to evaluate what factors correlate to higher number of casualties in school shootings, in hopes that this analysis will help lead to identification of good preventive measures. In addition, we will also be looking at whether our data is in line with a study (Vossekuil et al., 2002) that analyzed 37 school shooting which identified attackers as all male, 95% current student of the school, 81% single attackers. As researchers claim that most firearms in school shootings were acquired from a family member (Vossekuil et al., 2002; Levin & Madfis, 2009), this analysis will also explore whether such findings apply to the data we have. 

Throughout this project, our team will look into the ```schools-shootings-data.csv``` as provided by the Washington Post. The data consists of any act of gunfire at schools (primary, secondary) that occurred during school hours, or immediately before or after hours, since the Columbine High massacre on April 20, 1999. As such, 238 incidences of school shootings have been included in the dataset. The data was collected by using “Nexis, news articles, open-source databases, law enforcement reports, information from school websites, and calls to schools and police departments”, as according to Washington Post. The variables included in this data are as follows: 

uid, nces_school_id, school_name, nces_district_id, district_name, date, school_year, year, time, day_of_week, city, state, school_type, enrollment, killed, injured, casualties, shooting_type, age_shooter1, gender_shooter1, race_ethnicity_shooter1, shooter_relationship1, shooter_deceased1, deceased_notes1, age_shooter2, gender_shooter2, race_ethnicity_shooter2, shooter_relationship2, shooter_deceased2, deceased_notes2, white, black, Hispanic, Asian, American_indian_alaska_native, Hawaiian_native_pacific_islander, two_or_more, resource_officer, weapon, weapon_source, lat, long, staffing, low_grade, high_grade, lunch, county, state_fips, county_fips, ulocale

More information on the description of these variables can be found under the codebook in the ```data``` folder. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Section 2. Data Analysis

```{r load-packages, message=FALSE}
library(tidyverse)
library(broom)
```

```{r load-data, message=FALSE}
school_shootings <- read_csv("../data/school-shootings-data.csv")
```

### In Line with Previous Literature?

##### Vossekuil's Study

```{r vossekuil, echo= FALSE}
literature <- school_shootings %>%
  mutate(gender_shooter1 = case_when(
    str_detect(gender_shooter1, "b") ~ "m",
    str_detect(gender_shooter1, "m") ~ "m",
    str_detect(gender_shooter1, "f") ~ "f"
  ))

literature %>%
  filter(gender_shooter1 != "NA") %>%
  count(gender_shooter1, gender_shooter2) %>%
  mutate(percentage = n/sum(n) * 100)

literature %>%
  filter(shooter_relationship1 != "NA") %>%
  count(shooter_relationship1) %>%
  mutate(percentage = n/sum(n) * 100)
```

Unlike the analysis presented in the study of Vossekuil et al. (2002) that indicated all attackers as male, our data shows that 4.5% of the first shooter were female, although there were no second shooters that were female. Out of 222 observations where enough information was given about the shooters, only 5 shootings had 2 shooters involved--which is only 2.25%, leaving 97.75% of the attacks as being committed by single attackers. This is a much bigger number compared to the 81% Vossekuil et al. reported in their study. Also, although Vossekuil reported that 95% of the attackers in their study were current students, in this data 125 out of 188 (66.5%) shootings were committed by current students, which is a smaller proprotion than that observed by Vossekuil.

##### Weapons from Family Members?

```{r weapon-resource}
literature %>%
  filter(is.na(weapon_source)) %>%
  nrow()

literature %>%
  filter(weapon_source != "NA") %>%
  mutate(wepsource_fam = case_when(
    str_detect(weapon_source, "brother") ~ TRUE,
    str_detect(weapon_source, "parents") ~ TRUE,
    str_detect(weapon_source, "family") ~ TRUE,
    str_detect(weapon_source, "father") ~ TRUE,
    str_detect(weapon_source, "father's") ~ TRUE,
    str_detect(weapon_source, "home") ~ TRUE,
    str_detect(weapon_source, "grandparents") ~ TRUE,
    str_detect(weapon_source, "grandfather") ~ TRUE,
    str_detect(weapon_source, "guardian") ~ TRUE,
    str_detect(weapon_source, "mother") ~ TRUE,
     str_detect(weapon_source, "parent") ~ TRUE,
     str_detect(weapon_source, "relative") ~ TRUE,
     str_detect(weapon_source, "relatives") ~ TRUE,
     str_detect(weapon_source, "stepfather") ~ TRUE,
     str_detect(weapon_source, "uncle") ~ TRUE
  )) %>%
  count(wepsource_fam) %>%
  mutate(percentage = n/sum(n) * 100)

literature %>%
  filter(shooter_relationship1 == "student") %>%
  filter(weapon_source != "NA") %>%
  mutate(wepsource_fam = case_when(
    str_detect(weapon_source, "brother") ~ TRUE,
    str_detect(weapon_source, "parents") ~ TRUE,
    str_detect(weapon_source, "family") ~ TRUE,
    str_detect(weapon_source, "father") ~ TRUE,
    str_detect(weapon_source, "father's") ~ TRUE,
    str_detect(weapon_source, "home") ~ TRUE,
    str_detect(weapon_source, "grandparents") ~ TRUE,
    str_detect(weapon_source, "grandfather") ~ TRUE,
    str_detect(weapon_source, "guardian") ~ TRUE,
    str_detect(weapon_source, "mother") ~ TRUE,
     str_detect(weapon_source, "parent") ~ TRUE,
     str_detect(weapon_source, "relative") ~ TRUE,
     str_detect(weapon_source, "relatives") ~ TRUE,
     str_detect(weapon_source, "stepfather") ~ TRUE,
     str_detect(weapon_source, "uncle") ~ TRUE
  )) %>%
  count(wepsource_fam) %>%
  mutate(percentage = n/sum(n) * 100)

```

Of 238 observations, 148 observations do not have the source of weapon included in the list. Of the remaining 90 observations, 50 were cases where the shooter gained access to weapons through family members (e.g. uncle, relatives, parents), which is 55.5% for the entire observations. When narrowing this down to shooters who were current students of the school, we get that 75.44% of the students get their weapons from their family members or home, which is in line with the literature that claim that most firearms in school shootings were acquired from a family member (Vossekuil et al., 2002; Levin & Madfis, 2009)

### Mutation of Variables

We begin by manipulating some of our variables in order to make them more useful. For example, many of our variables are raw and thus, converting them to percentages will be more useful and standardized. 
```{r mutation-races}
school_shootings <- school_shootings %>%
  mutate(white_percent = white/enrollment,
         black_percent = black/enrollment,
         hispanic_percent = hispanic/enrollment,
         asian_percent = asian/enrollment,
         american_indian_alaska_native_percent = american_indian_alaska_native/enrollment)
```

```{r lunch-percent}
school_shootings <- school_shootings %>%
  mutate(lunch_percent = lunch/enrollment)
```

```{r ulocale-factoring}
school_shootings <- school_shootings %>%
  mutate(locale_type = case_when(
    ulocale == 11 ~ "City",
    ulocale == 12 ~ "City",
    ulocale == 13 ~ "City",
    ulocale == 21 ~ "Suburb",
    ulocale == 22 ~ "Suburb",
    ulocale == 23 ~ "Suburb",
    ulocale == 31 ~ "Town",
    ulocale == 32 ~ "Town",
    ulocale == 33 ~ "Town",
    ulocale == 41 ~ "Rural",
    ulocale == 42 ~ "Rural",
    ulocale == 43 ~ "Rural"
  ))
```

```{r school-type, warning=FALSE}
school_shootings <- school_shootings %>%
  mutate(high_grade = case_when(high_grade == "PK" ~ 0,
TRUE ~ as.numeric(high_grade)),
low_grade = case_when( low_grade == "PK" ~ 0,
low_grade == "KG" ~ 0,
TRUE ~ as.numeric(low_grade)),
  )

school_shootings <- school_shootings %>%
  mutate(school_type = case_when(
  is.na(high_grade) ~ NA_character_,
  (high_grade >= 9) ~ "highschool",
  (high_grade >= 6) ~ "middleschool",
  TRUE ~ "elementary"
))
```

We create the following variables:

<center>

```white_percent```
```black_percent```
```hispanic_percent```
```asian_percent```
```american_indian_alaska_native_percent```
```school_type```
```lunch_percent```
```locale_type```

</center>

### Frequency of Annual School Shootings

```{r quan-shootings, fig.width=5, echo=FALSE}
school_shootings %>%
  group_by(year) %>%
  count() %>%
  ggplot(mapping = aes(x = year, y = n)) +
  geom_line() +
  labs(title = "Number of annual school shootings", subtitle = "1999 - 2018") +
  ylab("Frequency") +
  xlab("Year")
```

### Number of Casualties Across the Years

```{r freq-casualties, fig.width=6, echo=FALSE}
school_shootings %>%
  ggplot(mapping = aes(x = year, y = casualties)) +
  geom_point() +
  labs(title = "Casualties over the years 1999 - 2018") +
  ylab("Casualties") +
  xlab("Year")
```

### Distribution of Casualties

```{r dist-casualties, fig.width=5, echo=FALSE}
school_shootings %>%
  ggplot(mapping = aes(x = casualties)) +
  geom_bar() +
  labs(title = "Distribution of Casualties", subtitle = "School Shootings from 1999-2018") +
  xlab("Casualties") +
  ylab("Frequency")
```

### School Shootings by State
Here is a visualization of school shootings per state in the U.S.

```{r testing, message=FALSE, warning=FALSE}
library(mapdata)
library(maps)
library(reshape2)

theme_map <- function(base_size = 12, base_family = "") {
    # Starts with theme_grey and then modify some parts
    theme_bw(base_size = base_size, base_family = base_family) %+replace% theme(strip.background = element_blank(),
        strip.text.x = element_text(size = 18), strip.text.y = element_text(size = 18),
        axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_blank(), panel.background = element_blank(),
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), panel.margin = unit(0.1, "lines"),
        plot.background = element_blank(), plot.margin = unit(c(0, 0, 0, 0), "lines"), axis.line = element_blank(), legend.background = element_blank(),
        legend.margin = unit(0.1, "line"), legend.title = element_text(size = 14,
            colour = "black"), legend.text = element_text(size = 14, colour = "black",
            hjust = 0.2), legend.key = element_blank(), legend.key.width = unit(1.5,
            "line"), legend.key.height = unit(1.5, "line"))

}

school_shootings_state <- school_shootings %>%
  count(state) %>%
  mutate(Shootings = n) %>%
  select(state, Shootings)

missing_states <- data.frame("Shootings" = c(0,0,0,0,0,0,0,0), "state" = c("Delaware", "Idaho", "Iowa", "Maine", "North Dakota", "Vermont", "West Virginia", "Wyoming"), stringsAsFactors = FALSE)

school_shootings_state <- rbind(school_shootings_state, missing_states) %>% arrange(state) %>%
  mutate(state = tolower(state))

school_shootings_state$state <- as.factor(school_shootings_state$state)

school_shootings_state <- subset(school_shootings_state, state != "District of Columbia")

us <- map_data("state")

ggplot(school_shootings_state, aes(map_id = state)) + geom_map(aes(fill = Shootings), map = us) + 
    labs(title = "School Shooting Distribution in U.S.", fill = "Shootings") + expand_limits(x = us$long, y = us$lat) + 
    coord_map("polyconic") + theme_map() + scale_fill_gradient(low = "#FFCCCC", high = "#CB454A")
```

### Mean Casualties Estimation via Bootstrapping

# ```{r bootstrap}
# boot_dist <- school %>%
#   # specify the variable of interest
#   specify(response = rent) %>% 
#   # generate 15000 bootstrap samples
#   generate(reps = 15000, type = "bootstrap") %>% 
#   # calculate the median of each bootstrap sample
#   calculate(stat = "median")
# ```

### Multivariate Regression Analysis

Next, we eliminate variables that will not be useful in our model, such as those used for logistical purposes. 

```{r sel-vars}
shootings_var_rm <- school_shootings %>%
  select(school_name,
         date, 
         school_year, 
         year, 
         time, 
         day_of_week,
         lunch,
         city, 
         state, 
         school_type, 
         enrollment, 
         killed, 
         injured, 
         casualties, 
         shooting_type, 
         age_shooter1, 
         gender_shooter1, 
         race_ethnicity_shooter1, 
         shooter_relationship1, 
         shooter_deceased1,
         resource_officer, 
         weapon, 
         staffing, 
         white_percent, 
         black_percent, 
         hispanic_percent, 
         asian_percent, 
         lunch_percent,
         american_indian_alaska_native_percent,
         locale_type)
```

These are the variables that we will include in our model:

<center>
  
```enrollment``` 
```factor(shooting_type)```
```staffing```
```white_percent```
```black_percent```
```hispanic_percent```
```asian_percent```
```lunch_percent```
```factor(resource_officer)```
```factor(day_of_week)```
```american_indian_alaska_native_percent```

</center>

### Model Creation and Backwards Selection

We first create a full model, which includes all potential variables. The ```step()``` function automates the backwards selection process for us.

```casualties ~ enrollment + factor(shooting_type) + staffing + white_percent + black_percent + hispanic_percent + asian_percent + lunch_percent + factor(resource_officer) + factor(day_of_week)```

### Backwards Selection

```{r back-sel, results='hide'}
mod_cas <- lm(casualties ~ 
                enrollment +
                factor(shooting_type) +
                staffing +
                white_percent +
                black_percent +
                hispanic_percent +
                asian_percent +
                lunch_percent +
                factor(resource_officer) + 
                factor(day_of_week) +
                #factor(school_type) +
                #age_shooter1 +
                #factor(gender_shooter1)
                #factor(shooter_deceased1),
                american_indian_alaska_native_percent,
                #factor(locale_type),
              data = school_shootings)



#glance(mod_cas)$adj.r.squared

perf_model <- step(mod_cas, direction = "backward")

perf_model %>%
  tidy() %>%
  select(term, estimate)

glance(perf_model)$adj.r.squared

```

The perfect model is shown below:

```casualties ~ enrollment + factor(shooting_type) + staffing + 
    lunch_percent + factor(resource_officer) + american_indian_alaska_native_percent```

Its adjusted r-squared value is 0.1783479.



## Section 3. Discussion and Conclusion

Through this exploratory analysis, we identified some variables that correlate to a higher number of casualties for school shootings: the number of full time teachers at time of shooting (```staffing```), enrollment at school at time of shooting (```enrollment```), number of students eligible for free or reduced price lunch (```lunch```), type of shooting (```shooting_type```), and presence of resource officer or security guard at time of shooting (```resource_officer```). As correlation does not imply causation, we would not go as far as to say that these were the factors that caused more victims (e.g. the absence of a resource officer may not have directly caused higher number of casualties). Also, with an adjusted r-squared value of 0.169, only about 17% of variability in number of casualties can be explained by these 5 variables. 

Some possible explanations as to why the 5 variables were best predictors of number of casualties is as follows: the more full time teachers there were at the time of shooting, the more possible it would have been for teachers to collectively protect the students and follow emergency guidelines; more students being enrolled on school would lead to more potential victims; and indiscriminate shooting would have lead to more casualties than targeted shooting. The resource officer being present being correlated to more number of casualties is interesting and difficult to explain. Perhaps the officer being present may have somehow either triggered the students to commit more shootings out of fear, maybe the fact that the school had an officer was indicative of the school experiencing trouble with violence, or as indicated in the literature (Addington, 2009), having security measures in school may have lead to the school feeling like prison for the students, causing more emotional distress and resentment. The ```lunch``` variable is also interesting in that it shows how schools with a greater proportion of students receiving free lunch have fewer number of casualties during shootings. Although it is hard to explain why this is the case from the given data, one possibility is that these schools had better preventive measures established prior to the shooting occurrence.

A few limitations exist in this analysis. First, our data may have potentially been unreliable, as it got information from that available to the public--at times, the gender or age of the shooters wasn’t reported, for example. Also, there were several variables and data that would have been interesting to look at to provide a fuller understanding of our analysis such as "rate of reported bullying", as the literature suggests bullying as one of the main causes of school shootings. Although the ```lunch``` variable is indicative of the SES status of the students in the school, an "average/median family income" would have helped better explain the significance of this variable. Finally, because we didn't have any data for schools that don't have shootings, we were not able to analyze and assess how schools that experienced shootings were different from those that did not. More information on what type of preventive measures were in schools prior to the shooting would have also been interesting to see how effective these measures are (for our current data, we only have the variable ```resource_officer``` for this). If we were able to start over with the project, we would have addressed the aforementioned issues with the data. In terms of statistical analysis, although we looked at the number of casualties in a numerical perspective, we could have looked at it in a categorical way by comparing schools that had shootings with no casualties to schools that had shootings with one or more casualties. 
